\documentclass[11pt]{article}

\usepackage[T1]{fontenc}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{booktabs}
\usepackage[margin=1in]{geometry}

\title{Per-Anchor Uncertainty Formulas in VBOGS}
\author{VBOGS}
\date{}

\newcommand{\R}{\mathbb{R}}
\newcommand{\E}{\mathbb{E}}
\newcommand{\Dir}{\operatorname{Dirichlet}}
\newcommand{\Cat}{\operatorname{Categorical}}
\newcommand{\MVN}{\operatorname{MVN}}
\newcommand{\softmax}{\operatorname{softmax}}

\begin{document}
\maketitle

\section{Observed Colored Points}

Each stereo point is represented as a six-dimensional observation:

\[
  x_n^{\mathrm{w}}
  =
  \begin{bmatrix}
    p_n^{\mathrm{w}} \\
    c_n
  \end{bmatrix}
  \in \R^6,
  \qquad
  p_n^{\mathrm{w}} \in \R^3,
  \qquad
  c_n \in \R^3.
\]

Here \(p_n^{\mathrm{w}}\) is the world-frame position and \(c_n\) is RGB
color. VBOGS keeps both world-frame and normalized versions. Bucketing uses
world coordinates, while VBGS fitting uses globally normalized coordinates:

\[
  y_n
  =
  \frac{x_n^{\mathrm{w}} - o}{\sigma}
  =
  \begin{bmatrix}
    s_n \\
    c_n
  \end{bmatrix},
  \qquad
  s_n \in \R^3,
  \qquad
  c_n \in \R^3.
\]

The vector \(o\) is the global offset and \(\sigma\) is the global per-channel
standard deviation. The spatial part \(s_n\) is normalized xyz; the color part
\(c_n\) is normalized RGB.

\section{Point-to-Anchor Bucketing}

For an Octree-AnyGS anchor \(a\) at level \(\ell_a\), the cell size is

\[
  cell\_size_{\ell_a}
  =
  \frac{\mathrm{voxel\_size}}{\mathrm{fork}^{\ell_a}}.
\]

The grid coordinate of a world-space point at level \(\ell\) is

\[
  grid\_coord(p_n^{\mathrm{w}},\ell)
  =
  \operatorname{round}
  \left(
    \frac{p_n^{\mathrm{w}} - \mathrm{init\_pos}}{cell\_size_\ell}
  \right).
\]

A point \(n\) is assigned to anchor \(a\) when they belong to the same voxel/grid coordinate

\[
  grid\_coord(p_n^{\mathrm{w}}, \ell_a) = grid\_coord(a_{\mathrm{pos}}, \ell_a).
\]

The assignment is performed across all Octree-AnyGS levels, not only the
finest level. The point set for anchor \(a\) is

\[
  \mathcal{T}_a
  =
  \left\{
    n
    \;:\;
    grid\_coord(p_n^{\mathrm{world}}, \ell_a) = grid\_coord(a_{\mathrm{pos}}, \ell_a).
  \right\}.
\]

Only anchors with enough assigned points are fitted:

\[
  |\mathcal{T}_a|
  \geq
  \mathrm{MIN\_POINTS\_PER\_ANCHOR}.
\]

Anchors below this threshold are treated as unobserved and receive \(U_{\max}\)
later.

\section{Per-Anchor VBGS Mixture}

For each observed anchor \(a\), VBOGS fits a VBGS mixture to the normalized
points

\[
  Y_a = \{y_n : n \in \mathcal{T}_a\}.
\]

The mixture has \(K_a\) active components. For point \(n\) and component \(k\),
the latent component assignment is

\[
  z_n \sim \Cat(\pi_a),
  \qquad
  \pi_a \sim \Dir(\alpha_a).
\]

The spatial and color parts are modeled by component-specific Gaussian
posteriors:

\[
  s_n \mid z_n = k
  \sim
  \MVN(\mu^s_{ak}, (\Lambda^s_{ak})^{-1}),
\]

\[
  c_n \mid z_n = k
  \sim
  \MVN(\mu^c_{ak}, (\Lambda^c_{ak})^{-1}).
\]

In implementation terms:

\[
  s_n = y_n[0:3],
  \qquad
  c_n = y_n[3:6].
\]

\section{Responsibilities}

During fitting, each component receives a soft responsibility for each point.
The implemented score combines spatial likelihood, color likelihood, and the
mixture prior term:

\[
  \ell_{nk}
  =
  \E_q[\log p(s_n \mid k)]
  +
  \E_q[\log p(c_n \mid k)]
  +
  \log \bar{\pi}_{ak}.
\]

The code obtains these three terms as

\[
  \texttt{space\_logprob}
  +
  \texttt{color\_logprob}
  +
  \texttt{prior\_logprob}.
\]

Responsibilities are the softmax over components:

\[
  \gamma_{nk}
  =
  q(z_n = k)
  =
  \frac{\exp(\ell_{nk})}
       {\sum_{j=1}^{K_a} \exp(\ell_{nj})}.
\]

These responsibilities are converted into sufficient statistics:

\[
  N_{ak} = \sum_{n \in \mathcal{T}_a} \gamma_{nk},
\]

plus spatial and color first/second-order statistics. 

The responsibilities define weighted sufficient statistics for each component.
For component \(k\), the effective point count is

\[
  N_{ak}
  =
  \sum_{n \in \mathcal{I}_a}
  \gamma_{nk}.
\]

The responsibility-weighted spatial mean is

\[
  \bar{s}_{ak}
  =
  \frac{1}{N_{ak}}
  \sum_{n \in \mathcal{I}_a}
  \gamma_{nk} s_n.
\]

The responsibility-weighted color mean is

\[
  \bar{c}_{ak}
  =
  \frac{1}{N_{ak}}
  \sum_{n \in \mathcal{I}_a}
  \gamma_{nk} c_n.
\]

Thus, points with large \(\gamma_{nk}\) strongly influence component \(k\)'s
mean, while points with small \(\gamma_{nk}\) contribute only weakly.

The corresponding weighted spatial scatter statistic is

\[
  S^s_{ak}
  =
  \sum_{n \in \mathcal{I}_a}
  \gamma_{nk}
  \left(s_n - \bar{s}_{ak}\right)
  \left(s_n - \bar{s}_{ak}\right)^{T}.
\]

Similarly, the weighted color scatter statistic is

\[
  S^c_{ak}
  =
  \sum_{n \in \mathcal{I}_a}
  \gamma_{nk}
  \left(c_n - \bar{c}_{ak}\right)
  \left(c_n - \bar{c}_{ak}\right)^{T}.
\]

\section{Choosing the Number of Components}

The fit starts with \(K_{\mathrm{init}}\). If the anchor has enough points for
growth, VBOGS tries a larger component count and compares mean ELBO per point:

\[
  \bar{\mathcal{L}}_K
  =
  \frac{1}{|\mathcal{T}_a|}
  \sum_{n \in \mathcal{T}_a}
  \mathcal{L}_{n,K}.
\]

A larger \(K\) is accepted only if

\[
  \bar{\mathcal{L}}_{K_{\mathrm{next}}}
  -
  \bar{\mathcal{L}}_K
  \geq
  \mathrm{ELBO\_IMPROVEMENT\_TOL}.
\]

The final accepted component count is \(K_a\). The implementation stores this
as \(\texttt{final\_k}\).

\section{Mixture Weights}

After fitting, the expected mixture weight for active component \(k\) is
computed from the Dirichlet parameters:

\[
  w_{ak}
  =
  \E_q[\pi_{ak}]
  =
  \frac{\alpha_{ak}}
       {\sum_{j=1}^{K_a} \alpha_{aj}}.
\]

The weights satisfy

\[
  \sum_{k=1}^{K_a} w_{ak} = 1.
\]

They describe how much of anchor \(a\)'s observed point set is explained by
each component. They are VBGS mixture weights, not Octree-AnyGS opacity values.

\section{Spatial Posterior Entropy}

For each active component, the spatial posterior is represented by a
Normal-Wishart distribution over the unknown component mean and precision:

\[
  q(\mu^s_{ak}, \Lambda^s_{ak})
  =
  q(\Lambda^s_{ak})
  q(\mu^s_{ak} \mid \Lambda^s_{ak}),
\]

\[
  \Lambda^s_{ak}
  \sim
  \operatorname{Wishart}(U^s_{ak}, \nu^s_{ak}),
  \qquad
  \mu^s_{ak} \mid \Lambda^s_{ak}
  \sim
  \MVN
  \left(
    m^s_{ak},
    (\kappa^s_{ak}\Lambda^s_{ak})^{-1}
  \right).
\]

This entropy is therefore the uncertainty in the fitted component's spatial
parameters, not the predictive entropy of a new point \(s_n\). The component
mean \(m^s_{ak}\) does not appear in the entropy because translating a
Gaussian changes its location but not its volume. For each active component,
the spatial posterior parameters used by the entropy are

\[
  \kappa^s_{ak},
  \qquad
  U^s_{ak},
  \qquad
  \nu^s_{ak}.
\]

These correspond to the implementation fields
\(\texttt{spatial\_kappa}\), \(\texttt{spatial\_u}\), and
\(\texttt{spatial\_n}\). The symbol \(U^s_{ak}\) here is the Wishart scale
matrix; it is unrelated to the final anchor scalar \(U_a\).

Let the spatial dimension be \(d_s = 3\). Since the mean covariance depends on
the random precision \(\Lambda^s_{ak}\), the entropy needs the expected log
determinant of that precision. For the Wishart parameterization used by VBGS,
this expectation is

\[
  \E_q[\log |\Lambda^s_{ak}|]
  =
  d_s \log 2
  +
  \log |U^s_{ak}|
  +
  \psi_{d_s}
  \left(
    \frac{\nu^s_{ak}}{2}
  \right),
\]

where the multivariate digamma expands the log-determinant expectation over
the \(d_s\) precision dimensions:

\[
  \psi_d(x)
  =
  \sum_{i=1}^{d}
  \psi
  \left(
    x - \frac{i-1}{2}
  \right).
\]

The Wishart normalizer also introduces the multivariate log-gamma:

\[
  \log \Gamma_d(x)
  =
  \frac{d(d-1)}{4}\log \pi
  +
  \sum_{i=1}^{d}
  \log \Gamma
  \left(
    x - \frac{i-1}{2}
  \right).
\]

The implementation then decomposes the joint entropy with the chain rule:

\[
  H(\mu^s_{ak}, \Lambda^s_{ak})
  =
  H(\Lambda^s_{ak})
  +
  \E_{q(\Lambda^s_{ak})}
  \left[
    H(\mu^s_{ak} \mid \Lambda^s_{ak})
  \right].
\]

The first term measures uncertainty in the component precision/covariance. The
second term measures uncertainty in the component mean after averaging over the
precision posterior. Expanding those two terms gives the spatial
Normal-Wishart entropy used by the implementation:

\[
\begin{aligned}
  H^s_{ak}
  &=
  H_{\mathrm{Wishart}}(\Lambda^s_{ak})
  +
  H_{\mathrm{Normal}}(\mu^s_{ak} \mid \Lambda^s_{ak}) \\
  &=
  \left[
    -\frac{1}{2}(\nu^s_{ak} - d_s - 1)
      \E_q[\log |\Lambda^s_{ak}|]
    + \frac{1}{2}\nu^s_{ak} d_s
    + \frac{1}{2}\nu^s_{ak} d_s \log 2
    + \frac{1}{2}\nu^s_{ak}\log |U^s_{ak}|
    + \log \Gamma_{d_s}
      \left(
        \frac{\nu^s_{ak}}{2}
      \right)
  \right] \\
  &\quad+
  \left[
    \frac{1}{2}d_s(1 + \log 2\pi)
    - \frac{1}{2}d_s\log \kappa^s_{ak}
    - \frac{1}{2}\E_q[\log |\Lambda^s_{ak}|]
  \right].
\end{aligned}
\]

The normal part has the same form as a Gaussian entropy, except the precision
log determinant is replaced by its posterior expectation. The term
\(-\frac{1}{2}d_s\log \kappa^s_{ak}\) is why a larger \(\kappa^s_{ak}\)
directly lowers spatial entropy: it tightens the posterior over the component
mean. The Wishart part depends on both \(U^s_{ak}\) and \(\nu^s_{ak}\), so
\(\nu^s_{ak}\) should be read together with the fitted scale matrix rather
than as a standalone uncertainty score. In practice, more assigned points tend
to increase \(\kappa^s_{ak}\) and \(\nu^s_{ak}\), and the posterior typically
contracts when the points agree on a compact spatial structure.

All of these quantities are computed in globally normalized xyz coordinates.
That is important because differential entropy changes under rescaling: a
per-anchor normalization would make \(H^s_{ak}\) values incomparable across
anchors unless the Jacobian correction were added back.

\section{Color/Delta Posterior Entropy}

For the color/delta posterior, the implementation uses

\[
  \kappa^c_{ak},
  \qquad
  U^c_{ak},
  \qquad
  \nu^c_{ak},
\]

stored as \(\texttt{delta\_kappa}\), \(\texttt{delta\_u}\), and
\(\texttt{delta\_n}\). Let the color dimension be \(d_c = 3\). The code uses

\[
  \E_q[\Lambda^c_{ak}]
  =
  \nu^c_{ak} U^c_{ak}.
\]

The covariance for the posterior mean is therefore

\[
  \Sigma^c_{ak}
  =
  \left(
    \kappa^c_{ak}\nu^c_{ak}U^c_{ak}
  \right)^{-1}.
\]

Equivalently, the implemented log determinant is

\[
  \log |\Sigma^c_{ak}|
  =
  -\log |U^c_{ak}|
  -
  d_c\log \nu^c_{ak}
  -
  d_c\log \kappa^c_{ak}.
\]

The color/delta entropy is the multivariate Gaussian entropy:

\[
  H^c_{ak}
  =
  \frac{1}{2}
  \left[
    d_c(1 + \log 2\pi)
    +
    \log |\Sigma^c_{ak}|
  \right].
\]

\section{Component Entropy and Anchor Uncertainty}

The entropy for component \(k\) in anchor \(a\) is

\[
  H_{ak}
  =
  H^s_{ak}
  +
  H^c_{ak}.
\]

The final observed-anchor uncertainty is the mixture-weighted per-component
posterior entropy:

\[
  U_a
  =
  \sum_{k=1}^{K_a}
  w_{ak} H_{ak}.
\]

This is the scalar written into \(\texttt{U.npy}\) for observed anchors whose
fit completed.

\section{Unobserved Anchors}

If an anchor does not have enough points, or if its fit did not complete, the
final uncertainty array receives

\[
  U_a = U_{\max}.
\]

By default,

\[
  U_{\max}
  =
  \max_{b \in \mathcal{O}} U_b,
\]

where \(\mathcal{O}\) is the set of anchors with finite observed uncertainty.
The user may override this with \(\texttt{--u-max}\).

\section{Dirichlet Entropy Diagnostic}

The implementation also computes Dirichlet entropy for diagnostics:

\[
  H_{\Dir}(\alpha_a)
  =
  \log B(\alpha_a)
  +
  (\alpha_{a0} - K_a)\psi(\alpha_{a0})
  -
  \sum_{k=1}^{K_a}
  (\alpha_{ak} - 1)\psi(\alpha_{ak}),
\]

where

\[
  \alpha_{a0}
  =
  \sum_{k=1}^{K_a}
  \alpha_{ak},
  \qquad
  \log B(\alpha_a)
  =
  \sum_{k=1}^{K_a}\log\Gamma(\alpha_{ak})
  -
  \log\Gamma(\alpha_{a0}).
\]

This value is saved in \(\texttt{uncertainty\_components.npz}\), but it is not
added to \(U_a\). The current VBOGS definition is

\[
  \boxed{
    U_a
    =
    \sum_{k=1}^{K_a}
    \frac{\alpha_{ak}}
         {\sum_j \alpha_{aj}}
    \left(
      H^s_{ak} + H^c_{ak}
    \right)
  }.
\]

\section{Implementation Mapping}

\begin{center}
\begin{tabular}{ll}
\toprule
Formula quantity & Implementation field or function \\
\midrule
\(\alpha_{ak}\) & \(\texttt{alpha}\) \\
\(K_a\) & \(\texttt{final\_k}\) \\
\(w_{ak}\) & \(\texttt{alpha / sum(alpha)}\) over active components \\
\(\kappa^s_{ak}\) & \(\texttt{spatial\_kappa}\) \\
\(U^s_{ak}\) & \(\texttt{spatial\_u}\) \\
\(\nu^s_{ak}\) & \(\texttt{spatial\_n}\) \\
\(\kappa^c_{ak}\) & \(\texttt{delta\_kappa}\) \\
\(U^c_{ak}\) & \(\texttt{delta\_u}\) \\
\(\nu^c_{ak}\) & \(\texttt{delta\_n}\) \\
\(H^s_{ak}\) & \(\texttt{normal\_wishart\_entropy(...)}\) \\
\(H^c_{ak}\) & \(\texttt{delta\_mvn\_entropy(...)}\) \\
\(U_a\) & \(\texttt{uncertainty[anchor\_id]}\) in \(\texttt{U.npy}\) \\
\bottomrule
\end{tabular}
\end{center}

\end{document}
